---
title: "Working with Genomic Ranges"
author: "Mireia Ramos-Rodriguez"
date: "March 14, 2019"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = FALSE)
```

# Introduction to GenomicRanges^[Reference card extracted from [mikelove/bioc-refcard](https://github.com/mikelove/bioc-refcard)]

The [GenomicRanges](http://bioconductor.org/packages/release/bioc/html/GenomicRanges.html) serves as the foundation for representing genomic locations within the Bioconductor project. This package lays a foundation for genomic analysis by introducing three classes (GRanges, GPos, and GRangesList), which are used to represent genomic ranges, genomic positions, and groups of genomic ranges. 

```{r}
library(GenomicRanges)
z <- GRanges("chr1",IRanges(1000001,1001000),strand="+")
start(z)
end(z)
width(z)
strand(z)
mcols(z) # the 'metadata columns', any information stored alongside each range
ranges(z) # gives the IRanges
seqnames(z) # the chromosomes for each ranges
seqlevels(z) # the possible chromosomes
seqlengths(z) # the lengths for each chromosome
```

# Intra-range methods

Affects ranges independently

__function__ | __description__  
-------------- | ----------------------------------------------------------- 
`shift` | moves left/right  
`narrow` | narrows by relative position within range  
`resize` | resizes to width, fixing start for +, end for -  
`flank` | returns flanking ranges to the left +, or right -  
`promoters` | similar to flank  
`restrict` | restricts ranges to a start and end position  
`trim` | trims out of bound ranges  
`+/-` | expands/contracts by adding/subtracting fixed amount  
`*` | zooms in (positive) or out (negative) by multiples  

# Inter-range methods

Affects ranges as a group

__function__ | __description__
-------------- | -----------------------------------------------------------
`range` | one range, leftmost start to rightmost end
`reduce` | cover all positions with only one range
`gaps` | uncovered positions within range
`disjoin` | breaks into discrete ranges based on original starts/ends

# Nearest methods

Given two sets of ranges, `x` and `subject`, for each range in `x`, returns...

__function__    | __description__
-------------- | -----------------------------------------------------------
`nearest`   | index of the nearest neighbor range in subject  
`precede`  | index of the range in subject that is directly preceded by the range in x  
`follow`   | index of the range in subject that is directly followed by the range in x  
`distanceToNearest`  | distances to its nearest neighbor in subject (Hits object)  
`distance`   | distances to nearest neighbor (integer vector)  

A Hits object can be accessed with `queryHits`, `subjectHits` and `mcols` if a distance is associated.

# set methods

If `y` is a GRangesList, then use `punion`, etc. All functions have default `ignore.strand=FALSE`, so are strand specific.

```{r}
union(x,y) 
intersect(x,y)
setdiff(x,y)
```

# Overlaps

```{r}
x %over% y  # logical vector of which x overlaps any in y
fo <- findOverlaps(x,y) # returns a Hits object
queryHits(fo)   # which in x
subjectHits(fo) # which in y 
```

# Seqnames and seqlevels

[GenomicRanges](http://www.bioconductor.org/packages/release/bioc/html/GenomicRanges.html) and [GenomeInfoDb](http://www.bioconductor.org/packages/release/bioc/html/GenomeInfoDb.html)

```{r}
gr.sub <- gr[seqlevels(gr) == "chr1"]
seqlevelsStyle(x) <- "UCSC" # convert to 'chr1' style from "NCBI" style '1'
```
